---
title: "R Gotcha! - Episode 1"
author: "Dr. Nelson Roque"
format: html
editor: visual
---

```{r}
# Load libraries ----
library(tidyverse)  # For data wrangling and visualization
library(digest)     # For generating data hashes
```

# Problem: The Quirky Data Import Issue

When working with R, we assume that loading the same dataset using different functions should yield identical data structures. **But is that always the case?**

Let's explore a subtle yet important quirk in data loading that can affect your analyses without you even realizing it.

------------------------------------------------------------------------

## Loading Data: The Sneaky Difference

Here, we load the same CSV file using two different functions:

```{r}
# Load two datasets ----
df_a = read.csv("../data/sample_data_with_header.csv")  # Base R function

df_b = readr::read_csv("../data/sample_data_with_header.csv")  # Tidyverse function
```

At first glance, these datasets may appear identical. But let’s do a quick comparison to check for hidden differences.

------------------------------------------------------------------------

## Problem, Visualized

To highlight potential differences, let’s extract the first column from each dataset and place them side by side:

```{r}
# Create a tibble to compare first columns ----
quick_compare = tibble::tibble(
  df_a_col_sample = df_a[,1],
  df_b_col_sample = df_b[,1] %>% pull(.)
)

# Print the tibble to see differences
quick_compare
```

### What’s Going On?

-   Base R’s `read.csv()` **automatically converts character columns into factors** unless specified otherwise.
-   `readr::read_csv()` **always treats text as character by default**.

This subtle difference can cause unexpected behavior in categorical data, leading to issues in factor-based operations.

------------------------------------------------------------------------

## Data Quality Checks: Can You Spot the Issue?

If we check for missing values, both datasets might appear the same:

```{r}
# Get missingness for each column ----
missing_a = colMeans(is.na(df_a))
missing_b = colMeans(is.na(df_b))
```

This is misleading because missingness checks don’t reveal structural differences in data types.

------------------------------------------------------------------------

## The Ultimate Check: Are These Dataframes Truly Identical?

A foolproof way to verify differences is by **hashing** each dataset and comparing them - generate unique hashes (e.g., MD5, SHA256):

```{r}
# Generate unique hashes for each dataset ----
df_a_hash = digest::digest(df_a)
df_b_hash = digest::digest(df_b)

# Compare hashes ----
df_a_hash == df_b_hash  
# If FALSE, the datasets are different!
```

------------------------------------------------------------------------

### The Reveal: What’s Different?

If the hashes don’t match, it confirms that `df_a` and `df_b` are not identical—even though they originated from the same file.

**Common Differences Include:** - Factor vs. Character Columns

```{r}
df_a_glimpse = dplyr::glimpse(df_a)
df_b_glimpse = dplyr::glimpse(df_b)

```

Definitely something funky when you stare with your eyeballs.

-   Different Default Column Types

```{r}
df_a_col_types = df_a %>%
  summarise(across(everything(), ~ class(.)[1], .names = "{.col}")) %>%
  pivot_longer(everything(), names_to = "col_name", values_to = "col_type")
  
df_b_col_types = df_b %>%
  summarise(across(everything(), ~ class(.)[1], .names = "{.col}")) %>%
  pivot_longer(everything(), names_to = "col_name", values_to = "col_type")
```

Hmm... something here.

-   Header Parsing Differences

```{r}
df_a_header = names(df_a)
df_b_header = names(df_b)

df_a_header == df_b_header
```

so that's not it...

------------------------------------------------------------------------

## Takeaways: How to Avoid This Gotcha

✅ Always inspect your data types using `str(df)` or `glimpse(df)` after loading.

✅ Use `stringsAsFactors = FALSE` in `read.csv()` if you want character data instead of factors.

✅ Prefer `readr::read_csv()` for a more predictable and consistent import behavior, but know the defaults.

✅ Hashing your data is a great way to check for hidden differences.

------------------------------------------------------------------------

## Acknowledgements

HUGE thank you to Dr. Daniel Roberts for flagging this issue for this episode of `rgotcha`.

------------------------------------------------------------------------

**Stay tuned for more moments when R Gotcha!** 🚀

📌 *Got an R gotcha you’d like us to cover? [Open an issue](https://github.com/nelsonroque/rgotcha/issues) and suggest a topic!*
