---
title: "R Gotcha! - Episode 2"
author: "Dr. Nelson Roque"
format: html
editor: visual
---

```{r}
# Load libraries ----
library(tidyverse)
library(stringr)
```

# Problem: The Leading Zero Disappearing Act

When working with R, we often assume that numbers behave like numbers and text behaves like text.

**But what happens when something looks numeric… but isn’t supposed to be?**

Let’s talk about one of the most dangerous silent data corruption traps in R:

👉 Leading zeros.

------------------------------------------------------------------------

## The Setup: IDs That Look Numeric

Many datasets contain identifiers such as:

- Participant IDs  
- ZIP codes  
- Device serial numbers  
- Census or FIPS codes  
- Visit labels  

These fields often contain leading zeros.

Let’s simulate a very normal dataset:

```{r}
df <- tibble(
  participant_id = c("00123", "04567", "08901"),
  visit_score = c(10, 12, 9)
)

df
```

Looks harmless enough.

------------------------------------------------------------------------

## The Sneaky Pipeline Step

Now imagine we want to standardize column types during preprocessing.

```{r}
df_bad <- df %>%
  mutate(participant_id = as.integer(participant_id)) %>%
  mutate(participant_id = as.character(participant_id))

df_bad
```

### Wait…

Where did the zeros go?

------------------------------------------------------------------------

## What Just Happened?

R numeric types store **values**, not formatting.

Mathematically:

```
00123 == 123
```

From a data integrity perspective?

Absolutely not.

------------------------------------------------------------------------

## Problem, Visualized

Let’s compare the original IDs against the converted ones:

```{r}
compare_ids <- tibble(
  original = df$participant_id,
  converted = df_bad$participant_id
)

compare_ids
```

Definitely something funky when you stare with your eyeballs.

------------------------------------------------------------------------

## Why This Is Dangerous

Once the leading zeros are removed:

- Longitudinal joins may fail  
- Database lookups may mismatch  
- Geospatial merges may break  
- Duplicate IDs may be created  
- Original formatting is permanently lost  

And here’s the scary part…

👉 R will not warn you.

------------------------------------------------------------------------

## A Hidden Source: CSV Import Guessing

Sometimes the issue starts before your pipeline even runs.

```{r}
# Simulate CSV import behavior
temp <- readr::type_convert(df)
temp
```

Column type guessing can silently coerce identifiers into numeric types depending on data patterns.

------------------------------------------------------------------------

## Data Quality Checks: Can You Spot It?

Missingness checks won’t reveal formatting loss:

```{r}
colMeans(is.na(df))
colMeans(is.na(df_bad))
```

Both appear identical.

Sneaky.

------------------------------------------------------------------------

## The Ultimate Check: Did We Corrupt Our IDs?

Let’s use hashing again 👀

```{r}
library(digest)

hash_original <- digest::digest(df$participant_id)
hash_converted <- digest::digest(df_bad$participant_id)

hash_original == hash_converted
```

If `FALSE`, something changed.

------------------------------------------------------------------------

## A Bonus Gotcha: Octal Confusion

Historically, numbers starting with `0` signaled **octal notation** in some programming languages.

Example:

```
010
```

Equals:

```
8 (base 8)
```

Modern R rarely interprets data this way directly, but this legacy behavior explains why leading zeros can be risky when parsed numerically.

------------------------------------------------------------------------

## The Safe Pattern

### Treat Identifiers As Character Always

```{r}
df_safe <- df %>%
  mutate(participant_id = as.character(participant_id))
```

------------------------------------------------------------------------

### If Padding Is Required

```{r}
df_safe %>%
  mutate(participant_id = str_pad(participant_id, width = 5, pad = "0"))
```

------------------------------------------------------------------------

## Defensive Programming: Build Guard Rails

Here’s a small helper function:

```{r}
safe_id <- function(x, width = NULL) {

  x <- as.character(x)

  if (!is.null(width)) {
    x <- stringr::str_pad(x, width, pad = "0")
  }

  x
}
```

------------------------------------------------------------------------

## Real-World Pipeline Check

If you want to scan for possible leading zero loss:

```{r}
df %>%
  mutate(id_length = nchar(participant_id))
```

Unexpected length variation can signal corruption.

------------------------------------------------------------------------

## Takeaways: How to Avoid This Gotcha

✅ Identifiers are not numeric data  
✅ Avoid converting ID columns to integer  
✅ Explicitly specify column types during import  
✅ Use padding utilities when formatting matters  
✅ Hash or length-check identifiers during QC  

------------------------------------------------------------------------

## Acknowledgements

Shoutout to every analyst who has ever debugged a mysterious join failure caused by leading zeros disappearing into the void.

You are not alone.

------------------------------------------------------------------------

**Stay tuned for more moments when R Gotcha!** 🚀  

📌 *Got an R gotcha you’d like us to cover? Open an issue on GitHub and suggest a topic!*
